Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
Link added no more anonymization

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.

>>>>

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this
direction~\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/merge_requests/10310}},
% \footnote{[Link removed due to anonymization]}
we leave this as future work.
